### 2015-10-14 Manuscript/SW_early_life_telomeres.Rmd
Terry edits, intro and methods

Lewis G. Spurgin^a,b^,  Kat Bebbington^a^, Eleanor A. Fairfield^a^, Jan Komdeur^c^, Terry Burke^d^, Hannah, L. Dugdale^c,d^,  and David S. Richardson^a,e,^.

a. School of Biological Sciences, University of East Anglia, Norwich Research Park, NR4 7TJ, United Kingdom
b. Department of Zoology, Edward Grey Institute, University of Oxford, Oxford, UK
c. Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands
d. Department of Animal and Plant Sciences, NERC Biomolecular Analysis Facility, University of Sheffield, Sheffield, UK
e. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles

@@

###Abstract
Understanding the links between early-life experiences and later life survival is fundamental to our understanding of life-history evolution. Telomeres - protective caps on the ends of chromosomes - shorten in response to early-life stress, and early-life telomere shortening is correlated with reduced later-life survival. As telomeres are a biomarker of the impact of early-life costs, they may provide a link between early-life environmental variation and later-life fitness. We studied how telomere dynamics link spatiotemporal variation in early-life conditions to survival using a long-term study of the Seychelles warbler (*Acrocephalus sechellensis*). We show that early-life telomere dynamics vary markedly across cohorts, and that individuals born in breeding seasons with higher food availability have longer telomeres and reduced telomere shortening. Within cohorts, we found that telomere length in nestlings was negatively related to tarsus length (which likely represents nestling growth), and positively related to the number of helpers present in the natal territory. Finally, we found that longer telomeres and reduced telomere shortening rates in early life were associated with increased survival later in life. Our results show that telomeres may provide a link between early-life conditions and late-life survival in wild populations.

@@

###Introduction
Being subject to favourable environmental conditions during development and growth can confer fitness advantages later in life (so called 'silver spoon effects') [@Grafen1988; @Monaghan2008]. Adult fitness can be affected by a range of early-life experiences, including variation in habitat quality and resource availability [@Madsen2000; @VandePol2006; @Hayward2013], population density [@Nussey2007;@Douhard2013] and natural or anthropogenic environmental disturbance [@Reid2003; @Cartwright2014]. Understanding the causes and consequences of these early-life experiences is key to understanding many ecological and evolutionary  processes, including  patterns of natural and sexual selection,  population growth rates and even local extinction [@Coulson2001; @Roach2014]. Understanding silver spoon effects is therefore of central interest to ecologists, evolutionary biologists and conservationists.

While it is clear that silver spoon effects can occur, we have little understanding of how and under what conditions early-life environments will affect adult fitness. Because the phenotypic consequences of an adverse environment can depend on an individual's initial  condition, and genetic or epigenetic makeup [@Hoffman2000; @Richards2006], the later-life consequences of a good or poor start are expected to vary among individuals, populations and species. Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships between the early-life conditions that are measured and adult phenotypes. Indeed, the pervasiveness of silver spoon effects varies between species [@Drummond2011], cohorts [@Reid2003] and sexes [@Wilkin2009]. To better understand the later-life consequences of early-life experiences, we therefore need to understand, or at least be able to measure, how the environment differentially affects individuals within a population.

A key process that has been associated with adult survival is the rate at which telomeres are shortened. Telomeres are protective caps on the ends of chromosomes which shorten with age [@Monaghan2006], and in response to oxidative stress levels which can be elevated  due to environmental factors [@VonZglinicki2002]. When telomeres reach a critical shortness cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003]. These links between telomere length and senescence have inspired a great deal of recent research into telomere ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014]. While there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Barrett2011; @Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, retaining signatures of ecological stress that are otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].

There is evidence from a range of taxa that the greatest rate of telomere loss occurs in early life [e.g. @Frenck1998; @Haussmann2003], and that the extent of this telomere shortening is influenced by the conditions experienced early in life [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015]. Importantly, early-life telomere dynamics are related to both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and to other parameters such as cognition [@Nettle2015a]. However, the link between early-life conditions and late-life survival is limited to a few species, and thus poorly understood. Moreover, how early-life telomere dynamics vary over spatial and temporal scales is not known.

The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent model system for studying senescence in the wild [reviewed in @Hammers2015]. Due to the isolated nature of the study population and intensive field monitoring, we have excellent survival data and sampling material spanning many years (see methods, below). Ecological conditions and population density on Cousin are variable over space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a]. Seychelles warblers remain on their natal territories for at least six months, and variation in natal territory quality has been linked to changes in levels of oxidative stress [@VandeCrommenacker2011a]. However, neither natal territory quality nor natal local density appear to be linked to early-life or adult survival [@Brouwer2006; @Hammers2013]. Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordinates) in the natal territory has been associated with increased survival later in life [@Brouwer2012]. Lastly, we have an established protocol for assessing absolute telomere length in this species [@Barrett2012], and telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost [@Barrett2013a]. Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced early in life, and to assess the later life consequences of early-life conditions.

In this study we test the prediction that telomeres link early-life environmental variation to late-life fitness. Because telomere dynamics are expected to reflect individual-level variation in the costs of early-life experiences, they may allow a more sensitive analysis of the effects of early-life environmental variation than would be possible with a direct comparison of the early-life environment and survival. With this in mind, we predict that: i) individuals raised in good quality conditions will have longer telomeres, and ii) having longer telomeres in early life is associated with greater survival.

@@

#####Study species and sampling
The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013]. The population of *ca*. 320 birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Spurgin2014a]. This species has two breeding seasons, running from June-August (main breeding season) and December-February (minor breeding season), when the breeding females on each of the *ca*. 115 territories lay one or, rarely, two or three eggs [@Komdeur1991]. As a result of this low reproductive output, combined with a higher mortality rate among first-year birds [39%; @Brouwer2006], cohort sizes in the Seychelles warbler are small (<50).

@@

Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively), and age is confirmed on the basis of eye colour [@Komdeur1991a] and previous captures [@Richardson2003b]. A blood sample (ca 25 $\mu$l) is taken from each caught bird via brachial venipuncture, and stored in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube at room temperature.

@@

Telomere lengths were measured for `r nrow(FlSA)+nrow(chicks)+nrow(Loss)` samples. Of these, `r nrow(FlSA)+nrow(chicks)` were taken cross-sectionally from birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`; `r nrow(chicks)` were nestlings less than one month old (chicks), while the remaining `r nrow(FlSA)` were fledglings and subadults aged between two and twelve months (hereafter referred to as fledglings). For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with an additional sample taken within two years of the original catch.  We measured absolute telomere quantity per diploid genome for all samples using a quantitative PCR (qPCR) assay  [full details in @Barrett2012].

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. Our sampling regime covers `r max(juv$LayYear)-min(juv$LayYear)` years, and temporal variation in rainfall and food availability on Cousin over this period is expected to result in variation  in a range of social and environmental variables [@Brouwer2012]. This makes it difficult to disentangle how spatial and temporal processes differentially affect early-life telomere dynamics. To overcome this, we analysed how variation in early-life conditions was related to variation in early-life telomere length separately among and within breeding seasons.

First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated average telomere length among all birds born in a given (major or minor) breeding season (i.e. a cohort), excluding cohorts where we had samples from $\leq$ 5 birds (n = `r nrow(juvseason)` cohorts). We tested for a difference in telomere length among cohorts using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among cohorts is explained by temporal variation in insect abundance.

We then conducted individual-level analyses to test how spatially varying early-life environmental and social conditions influence telomere length within cohorts, using general linear models. For these analyses, both response and explanatory variables were standardised by mean-centering within cohorts, thus controlling for any between-cohort temporal variation in telomere length and the social and ecological environment. As explanatory variables we included age class (nestling or fledgling), tarsus length, sex, territory quality and the number of helping and non-helping subordinate birds present in the natal territory. We also included interactions between age class and all the other variables. We used model averaging to assess how these variables shape telomere length and dynamics in early life. We created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model averaged coefficients, confidence intervals, and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].

For individuals with longitudinal data (n = `r nrow(Loss)`) we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events. To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013]. Using this dataset we repeated the above analyses of among and within cohort telomere dynamics, replacing telomere length with telomere loss as the response variable.

Finally, we analysed the effects of telomere dynamics on survival, again by looking at variation both within and among cohorts. First, we tested the prediction that cohort-level telomere dynamics are associated with survival by comparing the mean telomere length and rate of loss of each cohort with the mean lifespan of all individuals in that cohort, using linear regression. We then used parametric survival analysis to test whether individual-level telomere length and rate of loss were related to to survival. Again, we mean-centred variables within cohorts to eliminate between-season temporal variation in telomeres and survival. The survival time distribution was chosen by comparing the fit of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data, by visually inspecting plots and by comparing AIC values. As explanatory variables of survival we included early-life telomere length/loss (we ran a separate analysis for each, using the cross-sectional and longitudinal data, respectively), plus the factors identified as being related to telomere length in the previous analysis. With this approach we aimed to assess the effect of early-life telomere length on survival while controlling for potentially confounding variables that may be correlated with telomere dynamics.

>>>>

Lewis G. Spurgin^1,2^,  Kat Bebbington^1^, Eleanor A. Fairfield^1^, Jan Komdeur^3^, Terry Burke^4^, Hannah, L. Dugdale^3,4^,  and David S. Richardson^1,5,^.

a. School of Biological Sciences, University of East Anglia, Norwich Research Park, NR4 7TJ, United Kingdom
b. Department of Zoology, Edward Grey Institute, University of Oxford, Oxford, UK
c. Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands
d. Department of Animal and Plant Sciences, University of Sheffield, Sheffield, UK
e. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles

@@

###Abstract
Understanding the links between early-life experiences and later-life survival (so-called "silver spoon effects") is fundamental to understanding life-history evolution. Telomeres – the protective caps at the ends of chromosomes – shorten  in response to stress, and telomere shortening in early life is correlated with reduced later-life survival. Thus, telomere dyamics may help us quantify individual variation in early-life costs and silver spoon effects. We tested how telomere dynamics are related to spatiotemporal variation in early-life conditions and later-life survival  using a long-term study of the Seychelles warbler (*Acrocephalus sechellensis*). We show that telomere lengths and rates of shortening vary across cohorts, and that individuals born in breeding seasons with higher food availability had longer telomeres and reduced telomere shortening. Within cohorts, we found that telomere length in nestlings was negatively related to tarsus length (which probably  represents nestling growth), and positively related to the number of helpers present in the natal territory. Finally, we found that increased survival later in life was associated with longer telomeres and reduced telomere shortening rates in early life. Our results show that telomeres can both measure variation in early-costs and predict late-life survival in wild populations.

@@

###Introduction
Exposure to favourable environmental conditions during development and growth can confer fitness advantages later in life (so called 'silver spoon effects') [@Grafen1988; @Monaghan2008]. Adult fitness can be affected by a range of early-life experiences, including the quality of the habitat and available resources [@Madsen2000; @VandePol2006; @Hayward2013], population density [@Nussey2007;@Douhard2013] and natural or anthropogenic environmental disturbance [@Reid2003; @Cartwright2014]. Understanding the causes and consequences of these early-life experiences is key to understanding many ecological and evolutionary  processes, including  patterns of natural and sexual selection,  population growth rates and even local extinction [@Coulson2001; @Roach2014]. Understanding silver spoon effects is therefore of central interest to ecologists, evolutionary biologists and conservationists.

While it is clear that silver spoon effects can occur, we have little understanding of how and under what conditions early-life environments will affect adult fitness. However , we do expect the later-life consequences of a good or poor start to vary among individuals, populations and species because the phenotypic consequences of an adverse environment can depend on an individual's initial condition, and genetic or epigenetic makeup [@Hoffman2000; @Richards2006]. Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships between the early-life conditions that are measured and the resulting adult phenotypes. Indeed, the pervasiveness of silver spoon effects varies between species [@Drummond2011], cohorts [@Reid2003] and sexes [@Wilkin2009]. We therefore need to understand, or at least be able to measure, how the environment affects individuals differentially within a population in order to elucidate the later-life consequences of early-life experience.

Telomeres - the protective caps on the ends of chromosomes - may provide a solution to this problem. Telomeres shorten with age [@Monaghan2006], and in response to oxidative stress, which can be elevated  due to environmental factors [@VonZglinicki2002]. When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003]. This association between senescence and telomere length has inspired a great deal of recent research into telomere ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014]. While there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Barrett2011; @Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].

There is evidence from a range of taxa that the greatest rate of telomere loss occurs in early life [e.g. @Frenck1998; @Haussmann2003], and that the extent of this telomere shortening is influenced by the conditions experienced during that period [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015]. Importantly, early-life telomere dynamics have been associated with both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and with other parameters such as cognition [@Nettle2015a]. However, to our knowledge no studies have simultaneously analysed how telomere dynamics, early-life conditions and late-life survival are all related in a natural setting. Moreover, how early-life telomere dynamics vary over spatial and temporal scales is not known.

The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent model system for studying senescence in the wild [reviewed in @Hammers2015]. Due to the isolated nature of the study population and intensive field monitoring, we have unusually comprehensive survival data and tissue samples spanning many years (see Methods, below). Ecological conditions and warbler population density on Cousin vary across space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a]. Seychelles warblers remain on their natal territories for at least six months, and variation in the oxidative stress experienced by individuals is associated with natal  territory quality [@VandeCrommenacker2011a]. However, neither early-life nor adult survival appear to be associated with natal territory quality or natal local density [@Brouwer2006; @Hammers2013]. Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordinates) in the natal territory is associated with increased survival later in life [@Brouwer2012]. Lastly, we have an established protocol for assessing absolute telomere length in this species [@Barrett2012], and telomere length predicts survival independently of age in adult Seychelles warblers, suggesting that telomeres act as a biomarker of cost [@Barrett2013a]. Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced early in life, and to assess the later-life consequences of early-life conditions.

In this study we test the prediction that telomeres link early-life environmental variation to late-life fitness. Because telomere dynamics are expected to reflect individual-level variation in the costs of early-life experiences, they may allow a more sensitive analysis of the effects of early-life environmental variation than would be possible with a direct comparison of how survival is affected by the early-life environment. With this in mind, we predict that: i) individuals raised in higher quality conditions will have longer telomeres and reduced telomere shortening, and ii) longer telomeres and lower rates of telomere shortening in early life are associated with greater survival.

@@

#####Study species and sampling
The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013]. The population of *ca*. 320 birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Spurgin2014a]. This species has two breeding seasons, running from June-August (main breeding season) and December-February (minor breeding season), when the breeding females on each of the *ca*. 115 territories lay one or, rarely, two or three eggs [@Komdeur1991]. As a result of this low reproductive output, combined with higher mortality in first-year birds [39%; @Brouwer2006], cohort sizes in the Seychelles warbler are small (<50).

@@

Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively), and age is confirmed on the basis of eye colour [@Komdeur1991a] and previous captures [@Richardson2003b]. A blood sample (*ca* 25 $\mu$l) is taken from each bird captured via brachial venipuncture, and stored at room temperature in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube.

@@

Telomere lengths were measured in `r nrow(FlSA)+nrow(chicks)+nrow(Loss)` samples. Of these, `r nrow(FlSA)+nrow(chicks)` were taken cross-sectionally from birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`; `r nrow(chicks)` were nestlings less than one month old (chicks), while the remaining `r nrow(FlSA)` were fledglings and subadults aged between two and twelve months (hereafter referred to as fledglings). For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with an additional sample taken within two years of the original catch.  We measured absolute telomere quantity per diploid genome for all samples using a quantitative PCR (qPCR) assay  [full details in @Barrett2012].

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. Our sampling regime covers `r max(juv$LayYear)-min(juv$LayYear)` years, and temporal variation in rainfall and food availability on Cousin Island over this period is expected to result in variation  in a range of social and environmental variables [@Brouwer2012]. This makes it difficult to disentangle how spatial and temporal processes differentially affect early-life telomere dynamics. To overcome this, we analysed how variation in early-life telomere length was related to variation in early-life conditions both among and within breeding seasons.

First, to analyse whether temporal variation in food availability drives population-level variation in telomere dynamics, we calculated mean telomere length among all birds born in a given (major or minor) breeding season (i.e. a cohort), excluding cohorts where the sample size $\leq$ 5 birds (n = `r nrow(juvseason)` cohorts). We tested for a difference in telomere length among cohorts using a one-way ANOVA, and then used linear regression to test whether variation in mean telomere length among cohorts is explained by temporal variation in insect abundance.

We then conducted individual-level analyses to test how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts, using general linear models. For these analyses, both response and explanatory variables were standardised by mean-centring within cohorts, thus controlling for any between-cohort temporal variation in telomere length and the social and ecological environment. As explanatory variables we included age class (nestling or fledgling), tarsus length, sex, territory quality and the number of helping and non-helping subordinate birds present in the natal territory. We also included interactions between age class and all the other variables. We used model averaging to assess how these variables shape telomere length and dynamics in early life. We created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].

For individuals with longitudinal data (*n* = `r nrow(Loss)`) we calculated a rate of telomere loss between the first-year sample and earliest adult sample by subtracting the adult telomere length from the early-life telomere length and dividing this difference by the length of time (in days) between sampling events. To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013]. Using this dataset we repeated the above analyses of among and within cohort telomere dynamics, replacing telomere length with telomere loss as the response variable.

Finally, we analysed the effects of telomere dynamics on survival, again by looking at variation both within and among cohorts. First, we tested the prediction that cohort-level telomere dynamics are associated with survival by comparing the mean lifespan of all individuals within each cohort with mean telomere length and rate of loss in that cohort, using linear regression. We then used parametric survival analysis to test whether survival was related to individual-level telomere length and rate of loss. Again, we mean-centred variables within cohorts to eliminate between-season temporal variation in telomeres and survival. The survival time distribution was chosen by comparing the fit of the observed survival data against a range of distributions (exponential, Weibull, log-normal, gamma, Gompertz and generalized F) by visually inspecting plots and comparing AIC values. As explanatory variables of survival, we included early-life telomere length/loss (we ran a separate analysis for each, using the cross-sectional and longitudinal data, respectively), plus the factors identified as being related to telomere length in the previous analysis. With this approach we aimed to assess the relationship between early-life telomere length and survival, while controlling for potentially confounding variables that might be correlated with telomere dynamics.
